EGGs are adhesive!

### 2025-05-18 calco/final.tex
passata

\noindent
Thus, $L$ and $R$ are the left- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet that are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, still ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can still be obtained~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is still under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.

@@

\parbox{11cm}{\begin{itemize}
	\item		\emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one aside, if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
		\item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable;
		\item \emph{closed under decomposition}, or \emph{left-cancellable}, if whenever $g$ and $g\circ f$ belong to $\mathcal{A}$, then $f\in \mathcal{A}$. 
\end{itemize}}\hfill
\parbox{1cm}{
\xymatrix{A \ar[r]^{f} \ar[d]_{m}& B \ar[d]^{n} \\ C \ar[r]_{g} & D}}

	\begin{definition}

	Let $\mathcal{A}\subseteq \mor(\X)$ be a class of arrows in a category $\X$ and consider the cube below on the right.

@@

\begin{definition}[\cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}]
	Let $\X$ be a category and $\mathcal{M}$ a subclass of
	$\mon(\X)$  including  all isos, closed under composition, decomposition,  and stable under pullbacks and pushouts.  The category  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
	\begin{enumerate}
		\item it has \emph{$\mathcal{M}$-pullbacks}, i.e.~pullbacks along arrows of $\mathcal{M}$;
		\item it has \emph{$\mathcal{M}$-pushouts}, i.e.~pushouts along arrows of $\mathcal{M}$;
		\item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
	\end{enumerate}
	%
	A category $\X$ is said to be \emph{strictly $\mathcal{M}$-adhesive}
	if $\mathcal{M}$-pushouts are Van Kampen. We write $m\colon X \mto Y$ to denote that an arrow $m\colon X\to Y$ belongs to $\mathcal{M}$.
\end{definition}

\begin{remark}
	\label{rem:salva} Our notion of $\mathcal{M}$-adhesive category corresponds to what in \cite{ehrig2006fundamentals} is called \emph{weak adhesive HLR category}, while a strict $\mathcal{M}$-adhesive categories corresponds to \emph{adhesive HLR} ones. Finally, 	\emph{Adhesivity} and \emph{quasiadhesivity} 
	\cite{lack2005adhesive,garner2012axioms} coincide with strict
	$\mon(\X) $-adhesivity and strict $\reg(\X)$-adhesivity,
	respectively. On the other hand
\end{remark}

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of (co-)slice and functor categories \cite{mac2013categories}, as well with respect to subcategories, as shown by the following theorems~\cite[Thm.~4.15]{ehrig2006fundamentals}, \cite[Prop.~3.5]{lack2005adhesive} and  \cite[Thm.~2.12]{CastelnovoGM24}.

\begin{proposition}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category. Then the following holds
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category, $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to \A$ one preserving $\mathcal{M}$-pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/\X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{5pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ that is stable under pullbacks and pushouts, contains all isos, and is closed under composition and decomposition.
	\end{enumerate} 
\end{proposition}

@@

We can state some useful properties of $\mathcal{M}$-adhesive category.
(see, for instance, \cite[Thm.~4.26]{ehrig2006fundamentals} or \cite[Fact 2.6]{azzi2019essence}). A proof is provided in \Cref{regmono-proof}.
\begin{restatable}{proposition}{regmono}\label{prop:regmono}
	Let $\X$ be an $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item every $\mathcal{M}$-pushout square is also a pullback;
		\item every arrow in $\mathcal{M}$ is a regular mono.
	\end{enumerate}
\end{restatable}
%  [Proof in ]

@@

\begin{restatable}{lemma}{epicol}\label{lemma:nat_trans_reg_epi_canonical_arrow_reg_epi}
    Let $F,G\colon \D\rightrightarrows \X$ be two diagrams, and suppose that $\X$ has all colimits of shape $\D$. Let $(X, \{x_d\}_{d \in \D})$ and $(Y, \{y_d\}_{d\in D})$ be the colimits of $F$ and $G$, respectively.  If $\phi\colon  F \to G$ is a natural transformation whose components are regular epis, then the arrow induced by $\phi$ from $X$ to $Y$ is a regular epi.
\end{restatable}
%[Proof in \Cref{epicol-proof}]
}
We conclude this section exploring some properties of kernel pairs in an $\mathcal{M}$-adhesive category. 
The results below appear to be original, and we give their proofs in \Cref{kpp-proof}.

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be arrows admitting kernel pairs and suppose that the solid part of the four squares below is given. 
	%
	If the leftmost square is commutative, then there is a a unique arrow $k_h\colon K_f \to K_g$ making the other three commutative.
	\[\xymatrix@R=15pt{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}& K_f  \ar@{.>}[r]^{k_h} \ar[d]_{(\pi^1_f,\pi^2_f)} & K_g \ar[d]^{(\pi^1_g, \pi^2_g)}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z & X\times Z \ar[r]_{h\times h}& X\times Z}\]

		Moreover, the following hold
		\begin{enumerate}
			\item if $h$ is mono then $k_h$ is mono too;
			\item if the leftmost square is a pullback, then also the central two are so;
			\item if $h$ is mono and the leftmost square is a pullback then the rightmost is a pullback too.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

@@

%\todo{A questo punto la proposizione seguente possiamo anche buttarla via se serve recuperare spazio}
\noindent
\parbox{7.5cm}{
\begin{restatable}{proposition}{mpo}\label{lem:mpo}
	Let $\X$ be a strict $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube aside the top is an $\mathcal{M}$-pushout and all the vertical faces are pullbacks. Then the right square is a pushout.
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar[rr]_{g} & & D }}

@@

\noindent
\parbox{7.5cm}{\begin{restatable}{lemma}{mps}\label{prop:kerset}
	Suppose that in $\Set$ the commuting cube in the diagram on the left is given, whose top face is a pushout, the left and bottom faces are pullbacks,  and $n\colon B\mto D$ is an injection. 
	Then the following hold
	\begin{enumerate}
		\item the right face of the cube is a pullback too;
		\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
	\end{enumerate}
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }}

@@

\Cref{prop:left,prop:com} allow us to deduce immediately the following.

@@

Sources and targets are represented by lines between dots and squares: the lines from the sources of an hyperedge comes from the left of the box, while the lines to the targets 

@@

\begin{remark}\label{rem:functor}
	It is worth to point out, as first noted in \cite{bonchi2022string}, that $\hyp$ is equivalent to a category of presheaves. 
	%
	Indeed, consider the category $\catname{H}$ in which the set of objects is given by $ (\mathbb{N}\times \mathbb{N}) \cup \{\bullet\}$ and arrows are given by the identities $\id{k,l}$, $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$. 
	%
	The functors $\catname{H}\to \Set$ corresponds exactly to hypergaphs: nodes correspond to the image of $\bullet$ while the set of hyperedges with source of length $k$ and target of length $l$ corresponds to the image of $(k,l)$ (see \cite{CastelnovoGM24}).
\end{remark}

@@

\Cref{cor:mono} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\Sigma}$ and a characterisation of monos in it.

@@

$\hyp_{\Sigma}$, has a forgetful functor $U_{\Sigma}\colon \hyp_{\Sigma}\to \X$ which sends $(h,k)\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ to $U_{\hyp}(\mathcal{H}$). Now, $U_{\hyp}(\mathcal{G}^{\Sigma})=1$ thus, for every set $X$ we can define $\Delta_{\Sigma}(X)\colon \Delta_{\hyp}(X)\to \mathcal{G}^{\Sigma}$  as $(?_{O_\Sigma}, !_{X})$. It is straightforward to see that in this way we get a left adjoint to $U_\Sigma$.

@@

\begin{remark}\label[remark]{rem:mono2}By \Cref{rem:label}, we know that if $\mathcal{G}$ is a term graph then $t_{\mathcal{G}}=v_1\circ \tau_{\mathcal{G}}$, where $v_1$ is the coprojection of $V_{\mathcal{G}}$ into $V^\star_{\mathcal{G}}$.  Notice that since $t_{\mathcal{G}}$ is a mono then $\tau_{\mathcal{G}}$ is a mono too.
\end{remark}

@@

\begin{remark}
	$\tg$ in general does not have terminal objects. 
	%Consider an algebraic signature in $\Set$. 
	Since $U_{\tg}$ preserves limits, if a terminal object exists it must have the singleton as set of nodes, therefore the set of hyperedges must be empty or a singleton. 
	Hence, for a counterexample, it suffices to take as signature the one given by two operations $a$ and $b$, both of arity $0$.
	$\tg$ is not an adhesive category, either. 
	In particular, as noted in e.g.~\cite{CastelnovoGM24}, 
	 it does not have pushouts along all monos. 
\end{remark}

@@

\begin{lemma}\label[lemma]{prop:push} Consider three term graphs $(l_0, !_{V_\mathcal{G}})\colon \mathcal{G}\to \mathcal{G}^{\Sigma}$, $(l_1, !_{V_\mathcal{H}})\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ and $(l_2, !_{V_\mathcal{K}})\colon \mathcal{K}\to \mathcal{G}^{\Sigma}$. Given $(f_1, g_1)\colon (l_0, !_{V_\mathcal{G}})\to (l_1, !_{V_\mathcal{H}})$, $(f_2, g_2)\colon (l_0, !_{V_\mathcal{G}})\to (l_2, !_{V_\mathcal{K}})$, if $(f_1, g_1)$ is a regular mono, then its pushout  $(p, !_{V_{\mathcal{P}}})\colon \mathcal{P}\to \mathcal{G}^{\Sigma}$ in $\hyp_{\Sigma}$ along $(f_2, g_2)$ is a term graph too.
\end{lemma}

@@

\section{Adding equivalences to hypergraphical structures}
\label{hypereq}
The use of hypergraphs equipped with an equivalence relation over their nodes has been argued as a convenient way to express concurrency in the DPO approach to rewriting~\cite{concur2006}
This section presents the framework by means of adhesive categories, including also its version for term graphs, as a stepping stone towards an analogous characterisation for e-graphs.

@@

\noindent
\parbox{11.4cm}{
\begin{restatable}{lemma}{comp}\label{prop:eqhyp_complete}
Let $F\colon \D \to \EqHyp$ be a diagram and $(E_D, V_D, Q_D, s_D, t_D, q_D)$ the image of an object $D$. Then the following hold
\begin{enumerate}
		\item $F$ has a colimit which is preserved by $K$;
	\item consider a cone $(L, \{l_D\}_{D\in \D})$ limiting  for $K \circ F$ and let $((E, V), \{(\pi^D_E, \pi^D_V)\}_{D\in \D})$ be one for $T\circ F$, then $F$ has a limit $(E, V, Q, s, t, q)$ and there is a mono $m\colon Q\mto L$ such that the diagram on the right commutes for every $D\in \D$.
\end{enumerate}
\end{restatable}}\hfill 
\parbox{3cm}{\xymatrix{V \ar@{>>}[d]_{q} \ar[r]^{\pi^D_V}& V_D \ar@{>>}[dd]^{q_D}\\Q \ar@{>.>}[d]_{m}&\\ L \ar[r]_{l_d} & Q_D}}

@@

A proof is in \Cref{proof:regmono}.
We have now all the ingredient necessary to study the adhesivity properties of $\EqHyp$.  As a first step we need to introduce a class of monos.

@@

                &\emptyset \ar@{>->}[dd]|\hole \ar@{>->}[rr] \ar@{>->}[dl] &&

@@

                \end{tikzpicture} \ar@{>->}[dd] \ar@{>->}[dl] \\

@@

                \ar@<2ex>@{>->}[dd]_(.3){d}\\&

@@

                \ar@{>->}[rr]|\hole \ar@{>->}[dl] && 

@@

                \ar@{>->}[dl]^{n} \\

@@

We are now ready to exploit the previous propertis to prove \Cref{prop:kerset}, as promised.

>>>>

\noindent
Thus, $L$ and $R$ are the left- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet that are not affected by the rule itself.
%
Should a category be adhesive, and the arrows of the rules monomorphisms, the presence of a match ensures that
if the pushout complement $C \to G$ exists then it is unique, hence a rewriting step can be deterministically performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can be recovered~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the theory is still under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness,
% and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.

@@

\vspace{-.25cm}
\parbox{11cm}{\begin{itemize}
	\item		\emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one aside, if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
		\item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable;
		\item \emph{closed under decomposition}, or \emph{left-cancellable}, if whenever $g$ and $g\circ f$ belong to $\mathcal{A}$, then $f\in \mathcal{A}$. 
\end{itemize}}\hfill
\parbox{1cm}{
\xymatrix{A \ar[r]^{f} \ar[d]_{m}& B \ar[d]^{n} \\ C \ar[r]_{g} & D}}

	\begin{definition}
	Let $\mathcal{A}\subseteq \mor(\X)$ be a class of arrows in a category $\X$ and consider the cube below on the right.

@@

\begin{definition}[\cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}]
	Let $\X$ be a category and $\mathcal{M}$ a subclass of
	$\mon(\X)$  including  all isos, closed under composition, decomposition,  and stable under pullbacks and pushouts.  The category  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
	\begin{enumerate}
		\item it has \emph{$\mathcal{M}$-pullbacks}, i.e.~pullbacks along arrows of $\mathcal{M}$;
		\item it has \emph{$\mathcal{M}$-pushouts}, i.e.~pushouts along arrows of $\mathcal{M}$;
		\item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
	\end{enumerate}
	%
	A category $\X$ is said to be \emph{strictly $\mathcal{M}$-adhesive}
	if $\mathcal{M}$-pushouts are Van Kampen. {\color{red} We write $m\colon X \mto Y$} to denote that an arrow $m\colon X\to Y$ belongs to $\mathcal{M}$.
\end{definition}

\begin{remark}
	\label{rem:salva} Our notion of $\mathcal{M}$-adhesive category corresponds to what in \cite{ehrig2006fundamentals} is called \emph{weak adhesive HLR category}, while a strict $\mathcal{M}$-adhesive categories corresponds to \emph{adhesive HLR} ones. Finally, 	\emph{adhesivity} and \emph{quasiadhesivity} 
	\cite{lack2005adhesive,garner2012axioms} coincide with strict
	$\mon(\X) $-adhesivity and strict $\reg(\X)$-adhesivity,
	respectively. %On the other hand
\end{remark}

$\mathcal{M}$-adhesivity is well-behaved with respect to  the construction of (co-)slice and functor categories \cite{mac2013categories}, as well with respect to subcategories, as shown by the following properties, taken from~\cite[Thm.~4.15]{ehrig2006fundamentals}, \cite[Prop.~3.5]{lack2005adhesive} and  \cite[Thm.~2.12]{CastelnovoGM24}.

\begin{proposition}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category, $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to \A$ one preserving $\mathcal{M}$-pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/\X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{25pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ that is stable under pullbacks and pushouts, contains all isos, and is closed under composition and decomposition.
	\end{enumerate} 
\end{proposition}

@@

We can state some useful properties of $\mathcal{M}$-adhesive category
(see, for instance, \cite[Thm.~4.26]{ehrig2006fundamentals} or \cite[Fact 2.6]{azzi2019essence}). A proof is provided in \Cref{regmono-proof}.
\begin{restatable}{proposition}{regmono}\label{prop:regmono}
	Let $\X$ be an $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item every $\mathcal{M}$-pushout square is also a pullback;
		\item every arrow in $\mathcal{M}$ is a regular mono.
	\end{enumerate}
\end{restatable}
%  [Proof in ]

@@

\begin{restatable}{lemma}{epicol}\label{lemma:nat_trans_reg_epi_canonical_arrow_reg_epi}
    Let $F,G\colon \D\rightrightarrows \X$ be two diagrams, and suppose that $\X$ has all colimits of shape $\D$. Let $(X, \{x_d\}_{d \in \D})$ and $(Y, \{y_d\}_{d\in D})$ be the colimits of $F$ and $G$, respectively.  If $\phi\colon  F \to G$ is a natural transformation whose components are regular epis, then the arrow induced by $\phi$ from $X$ to $Y$ is a regular epi.
\end{restatable}
%[Proof in \Cref{epicol-proof}]
}
We conclude this section exploring some properties of kernel pairs in an $\mathcal{M}$-adhesive category. 
The results below are simple, yet they appear to be original, and we give their proofs in \Cref{kpp-proof}.

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be arrows admitting kernel pairs and suppose that the solid part of the four squares below is given. 
	%
	If the leftmost square is commutative, then there is a unique arrow $k_h\colon K_f \to K_g$ making the other three commutative.
	\[\xymatrix@R=15pt{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}& K_f  \ar@{.>}[r]^{k_h} \ar[d]_{(\pi^1_f,\pi^2_f)} & K_g \ar[d]^{(\pi^1_g, \pi^2_g)}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z & X\times Z \ar[r]_{h\times h}& X\times Z}\]

		Moreover, the following hold
		\begin{enumerate}
			\item if $h$ is a mono then $k_h$ is a mono;
			\item if the leftmost square is a pullback then the central two are pullbacks;
			\item if $h$ is mono and the leftmost square is a pullback then the rightmost is a pullback.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

@@

%\todo{A questo punto la proposizione seguente possiamo anche buttarla via se serve recuperare spazio}
\noindent
\parbox{7.5cm}{
\begin{restatable}{proposition}{mpo}\label{lem:mpo}
	Let $\X$ be a strict $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube aside the top face is an $\mathcal{M}$-pushout and all the vertical faces are pullbacks. Then the right square is a pushout.
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar[rr]_{g} & & D }}

@@

\noindent
\parbox{7.5cm}{\begin{restatable}{lemma}{mps}\label{prop:kerset}
	Suppose that in $\Set$ the commuting cube in the diagram on the left is given, whose top face is a pushout, the left and bottom faces are pullbacks,  and $n\colon B\mto D$ is an injection. 
	Then the following hold
	\begin{enumerate}
		\item the right face of the cube is a pullback;
		\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
	\end{enumerate}
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }}

@@

{\color{red} \Cref{prop:left,prop:com}} allow us to deduce immediately the following.

@@

Sources and targets are represented by lines between dots and squares: the lines from the sources of a hyperedge comes from the left of the box, while the lines to the targets 

@@

\begin{remark}\label{rem:functor}
	It is worth to point out, as first noted in \cite{bonchi2022string}, that $\hyp$ is equivalent to a category of presheaves. 
	%
	Indeed, consider the category $\catname{H}$ in which the set of objects is given by $ (\mathbb{N}\times \mathbb{N}) \cup \{\bullet\}$ and arrows are given by the identities $\id{k,l}$, $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$. 
	%
	The functors $\catname{H}\to \Set$ corresponds exactly to hypergraphs: nodes correspond to the image of $\bullet$ while the set of hyperedges with source of length $k$ and target of length $l$ corresponds to the image of $(k,l)$ (see \cite{CastelnovoGM24}).
\end{remark}

@@

{\color{red} \Cref{cor:mono}} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\Sigma}$ and a characterisation of monos in it.

@@

$\hyp_{\Sigma}$ has a forgetful functor $U_{\Sigma}\colon \hyp_{\Sigma}\to \X$ which sends $(h,k)\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ to $U_{\hyp}(\mathcal{H}$). Now, since $U_{\hyp}(\mathcal{G}^{\Sigma})=1$ 
for every set $X$ we can define $\Delta_{\Sigma}(X)\colon \Delta_{\hyp}(X)\to \mathcal{G}^{\Sigma}$  as $(?_{O_\Sigma}, !_{X})$. It is straightforward to see that in this way we get a left adjoint to $U_\Sigma$.

@@

\begin{remark}\label[remark]{rem:mono2}By \Cref{rem:label}, we know that if $\mathcal{G}$ is a term graph then $t_{\mathcal{G}}=v_1\circ \tau_{\mathcal{G}}$, where $v_1$ is the coprojection of $V_{\mathcal{G}}$ into $V^\star_{\mathcal{G}}$.  Notice that since $t_{\mathcal{G}}$ is a mono then $\tau_{\mathcal{G}}$ is a mono.
\end{remark}

@@

\begin{remark}
	$\tg$ in general does not have terminal objects. 
	%Consider an algebraic signature in $\Set$. 
	Since $U_{\tg}$ preserves limits, if a terminal object exists it must have the singleton as set of nodes, therefore the set of hyperedges must be empty or a singleton. 
	Hence, for a counterexample, it suffices to take a signature with two operations $a$ and $b$, both of arity $0$.
	$\tg$ is not an adhesive category, either. 
	In particular, as noted in e.g.~\cite{CastelnovoGM24}, 
	 it does not have pushouts along all monos. 
\end{remark}

@@

\begin{lemma}\label[lemma]{prop:push} Consider three term graphs $(l_0, !_{V_\mathcal{G}})\colon \mathcal{G}\to \mathcal{G}^{\Sigma}$, $(l_1, !_{V_\mathcal{H}})\colon \mathcal{H}\to \mathcal{G}^{\Sigma}$ and $(l_2, !_{V_\mathcal{K}})\colon \mathcal{K}\to \mathcal{G}^{\Sigma}$. Given $(f_1, g_1)\colon (l_0, !_{V_\mathcal{G}})\to (l_1, !_{V_\mathcal{H}})$, $(f_2, g_2)\colon (l_0, !_{V_\mathcal{G}})\to (l_2, !_{V_\mathcal{K}})$, if $(f_1, g_1)$ is a regular mono, then its pushout  $(p, !_{V_{\mathcal{P}}})\colon \mathcal{P}\to \mathcal{G}^{\Sigma}$ in $\hyp_{\Sigma}$ along $(f_2, g_2)$ is a term graph.
\end{lemma}

@@

\section{Adding equivalences to hypergraphical structures}
\label{hypereq}
The use of hypergraphs equipped with an equivalence relation over their nodes has been argued as a convenient way to express concurrency in the DPO approach to rewriting~\cite{concur2006}.
This section presents the framework by means of adhesive categories, including also its version for term graphs, as a stepping stone towards an analogous characterisation for e-graphs.

@@

\noindent
\parbox{11.4cm}{
\begin{restatable}{lemma}{comp}\label{prop:eqhyp_complete}
Consider a diagram $F\colon \D \to \EqHyp$ and let $(E_D, V_D, Q_D, s_D, t_D, q_D)$ be the image of an object $D$. Then the following hold
\begin{enumerate}
		\item $F$ has a colimit, which is preserved by $K$;
	\item consider a cone $(L, \{l_D\}_{D\in \D})$ limiting  for $K \circ F$ and let $((E, V), \{(\pi^D_E, \pi^D_V)\}_{D\in \D})$ be one for $T\circ F$, then $F$ has a limit $(E, V, Q, s, t, q)$ and there is a {\color{red} mono $m\colon Q\mto L$} such that the diagram on the right commutes for every $D\in \D$.
\end{enumerate}
\end{restatable}}\hfill 
\parbox{3cm}{\xymatrix{V \ar@{>>}[d]_{q} \ar[r]^{\pi^D_V}& V_D \ar@{>>}[dd]^{q_D}\\Q \ar@{>.>}[d]_{m}&\\ L \ar[r]_{l_d} & Q_D}}

@@

A proof is in \Cref{proof:regmono}.
We have now all the ingredients to study the adhesivity properties of $\EqHyp$.  As a first step we need to introduce a class of monos.

@@

                &\emptyset \ar@{>->}[dd]|\hole \ar@{>->}[rr] \ar@{>->}@(dl,)[dl] &&

@@

                \end{tikzpicture} \ar@{>->}[dd] \ar@{>->}@(dl,)[dl]!<6ex,0ex> \\

@@

                \ar@<2ex>@{>->}[dd]\\&

@@

                \ar@{>->}[rr]|\hole \ar@{>->}@(dl,)[dl] && 

@@

                \ar@{>->}@(,r)[dl] \\

@@

We are now ready to exploit the previous properties to prove \Cref{prop:kerset}. % as promised.
